1 Report ASP20 Boost

### 2020-06-04 Reports/report_1/1_REPORT.tex
write first report

---
title: "Report ASP20 Boost"
subtitle: "First Report"
author:
- "Johannes Strauß"
- "Levin Wiebelt"
- "Sebastian Aristizabal"
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: [nothing, nothingness]
documentclass: report
bibliography: ["citavi_27.05.bib", "packages.bib"]
biblio-style: apalike
geometry: "left=4cm, right=3cm, top=2.5cm, bottom=2.5cm"
lof: FALSE
lot: FALSE
output: 
    pdf_document:
      toc: true
      toc_depth: 2
      latex_engine: xelatex
      keep_tex: yes
      template: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# 1. Introduction:

The course "Advanced Statistical Programming with R" consists in implementing a package to address location-scale regression. 
The sample dataset consists of a response y, that's expectation is in linear relationship to a set of predictors xi. 
Its variance, in contrast, is dependent on a different set of linear predictors zi, however transformed by the response function g(x) = exp(x). 
This ensures positiveness of the variances. 
The goal approach is to estimate the effects of both: xi on the expectation of y (location) - this is captured by the estimates beta, and zi on the variance of y (scale) - this is captured by the estimates gamma. Our student group `asp20boost` solves this task applying the concept of boosting.

@@

# 2. Concept: Boosting

The first step in our progress was to understand the theoretical concept of boosting, and which types of problems it is able to solve. In the following the term "booster" refers to a boosting algorithm.

## Location-Booster

To explain the concept and follow our process of understanding we first explain boosting for location parameters. Boosting scale parameters requires another way of thinking about it and constituted a first milestone in our understanding and implementation.
The boosting concept relies on the idea of iterative estimation. 
To estimate the location parameters beta via boosting, one starts with initial estimates, which may be far from optimal. 
However, with estimates and the response vector at hand, residuals can be calculated. 
The boosting-algortihm focuses on those calculated residuals.
The effect of the predictors xi on these residuals is estimated by least squares and the resulting effect is added to the location-estimate beta, adjusted by some learning rate. 
This yields a better fit of the location-model, which results in smaller residuals.
These new residuals are estimated again in the next iteration, yielding smaller effect sizes, and hence a convergence of the boosting algorithm towards the OLS-estimate.
A simple booster stops after a fixed number of iterations is reached.
**(comp. Kneib et al)**

## Scale-Booster

[Loss-function erklären: Boosting can be viewed as lowering iteratively the amount of a loss function, which is in our case the score of the normal distribution]

The concept for boosting scale parameters is the same. However, residuals for the variance estimates can not be calculated, since there is no such thing as "observed" variances for individual response values. The solution is to substitute the calculated residuals with a different term: the derivative of the loss function, which in our case is the log-likelihood of the normal distribution. Instead of estiamting stepwise residuals, one estiamtes hence stepwise the score.

@@

We employed the resid-function of the asp20model class, on which our class is built on.
<!-- The working and response residuals are the observed values -->
<!-- minus the fitted values for the location. -->
<!-- The deviance and Pearson residuals are weighted residuals, -->
<!-- divided by the fitted values for the scale. -->

# 3. Implementation

@@

The simple booster for gamma consists in the same two repeating steps of first estimating a term, and second updating the scale parameters gamma by adding the estimated effect - adjusted by a learning rate. 
The term to be estimated, however, is not the residual, but the derivative of the loglikelihood of the normal distribution. 
To calculate this, our code makes use of the resid-function(), but this time passing the argument "deviance", which results in residuals adjusted by the fitted scale estimates.

[Is the gamma booster dependent on simultaneous running of beta booster?]

@@

# 4. Functionality: Componentwise Boosting

@@

The implementation in our package works via the extension of the LocationScaleRegressionClass by two active fields "bestFittingVariableBeta" and "bestFittingVariableGamma". 
These functions partition the design matrix X, respectively Z, into its single columns and then estimate the residuals - respectively the score - seperately for each component, and determine loss functions for a hypothetical update with the respective component. 
Comparing with the old loss function value, the highest loss-improvement may be determined and the respective component is used to update the parameter-vector.

@@

* create public fields for the heavily used score-function values
* move the calculation of componentwise losses to an external function
* harmonize boosting and componentwise boosting into one external function, determining the mode of operation by an argument 'componentwise = TRUE'

@@

# 5. Prospects

## Further Functionalities

@@

## Stability

@@

## Performance

* A major performance problem in our code is that small learning rates for the beta-booster cause processing time of the code to increase sharply. This remains to be solved.

@@

## Further

* Allow user-input of the LocationScaleRegressionBoost-model in form of a dataframe, for example by allowing to pass an optional "data"-argument as known from the lm-call. This may reduce proneness to input-errors.

* Visualize the results of the boosted estimates

* Document the extension of the R6-Class properly and inherit documentation of the LocationScaleRegression-Class.

## Application

[Johannes: Munich Rent Data? - Identify Outliers]

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{report}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Report ASP20 Boost},
  pdfauthor={Johannes Strauß; Levin Wiebelt; Sebastian Aristizabal},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[left=4cm, right=3cm, top=2.5cm, bottom=2.5cm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{Report ASP20 Boost}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{First Report}
\author{Johannes Strauß \and Levin Wiebelt \and Sebastian Aristizabal}
\date{04 Juni 2020}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{introduction}{%
\chapter{1. Introduction:}\label{introduction}}

The course ``Advanced Statistical Programming with R'' consists in
implementing a package to address location-scale regression. The sample
dataset consists of a response y, that's expectation is in linear
relationship to a set of predictors xi. Its variance, in contrast, is
dependent on a different set of linear predictors zi, however
transformed by the response function g(x) = exp(x). This ensures
positiveness of the variances. The goal approach is to estimate the
effects of both: xi on the expectation of y (location) - this is
captured by the estimates beta, and zi on the variance of y (scale) -
this is captured by the estimates gamma. Our student group
\texttt{asp20boost} solves this task applying the concept of boosting.

@@

\hypertarget{concept-boosting}{%
\chapter{2. Concept: Boosting}\label{concept-boosting}}

The first step in our progress was to understand the theoretical concept
of boosting, and which types of problems it is able to solve. In the
following the term ``booster'' refers to a boosting algorithm.

\hypertarget{location-booster}{%
\section{Location-Booster}\label{location-booster}}

To explain the concept and follow our process of understanding we first
explain boosting for location parameters. Boosting scale parameters
requires another way of thinking about it and constituted a first
milestone in our understanding and implementation. The boosting concept
relies on the idea of iterative estimation. To estimate the location
parameters beta via boosting, one starts with initial estimates, which
may be far from optimal. However, with estimates and the response vector
at hand, residuals can be calculated. The boosting-algortihm focuses on
those calculated residuals. The effect of the predictors xi on these
residuals is estimated by least squares and the resulting effect is
added to the location-estimate beta, adjusted by some learning rate.
This yields a better fit of the location-model, which results in smaller
residuals. These new residuals are estimated again in the next
iteration, yielding smaller effect sizes, and hence a convergence of the
boosting algorithm towards the OLS-estimate. A simple booster stops
after a fixed number of iterations is reached. \textbf{(comp. Kneib et
al)}

\hypertarget{scale-booster}{%
\section{Scale-Booster}\label{scale-booster}}

{[}Loss-function erklären: Boosting can be viewed as lowering
iteratively the amount of a loss function, which is in our case the
score of the normal distribution{]}

The concept for boosting scale parameters is the same. However,
residuals for the variance estimates can not be calculated, since there
is no such thing as ``observed'' variances for individual response
values. The solution is to substitute the calculated residuals with a
different term: the derivative of the loss function, which in our case
is the log-likelihood of the normal distribution. Instead of estiamting
stepwise residuals, one estiamtes hence stepwise the score.

@@

We employed the resid-function of the asp20model class, on which our
class is built on.

\hypertarget{implementation}{%
\chapter{3. Implementation}\label{implementation}}

@@

The simple booster for gamma consists in the same two repeating steps of
first estimating a term, and second updating the scale parameters gamma
by adding the estimated effect - adjusted by a learning rate. The term
to be estimated, however, is not the residual, but the derivative of the
loglikelihood of the normal distribution. To calculate this, our code
makes use of the resid-function(), but this time passing the argument
``deviance'', which results in residuals adjusted by the fitted scale
estimates.

{[}Is the gamma booster dependent on simultaneous running of beta
booster?{]}

@@

\hypertarget{functionality-componentwise-boosting}{%
\chapter{4. Functionality: Componentwise
Boosting}\label{functionality-componentwise-boosting}}

@@

The implementation in our package works via the extension of the
LocationScaleRegressionClass by two active fields
``bestFittingVariableBeta'' and ``bestFittingVariableGamma''. These
functions partition the design matrix X, respectively Z, into its single
columns and then estimate the residuals - respectively the score -
seperately for each component, and determine loss functions for a
hypothetical update with the respective component. Comparing with the
old loss function value, the highest loss-improvement may be determined
and the respective component is used to update the parameter-vector.

@@

\begin{itemize}
\tightlist
\item
  create public fields for the heavily used score-function values
\item
  move the calculation of componentwise losses to an external function
\item
  harmonize boosting and componentwise boosting into one external
  function, determining the mode of operation by an argument
  `componentwise = TRUE'
\end{itemize}

@@

\hypertarget{prospects}{%
\chapter{5. Prospects}\label{prospects}}

\hypertarget{further-functionalities}{%
\section{Further Functionalities}\label{further-functionalities}}

@@

\hypertarget{stability}{%
\section{Stability}\label{stability}}

@@

\hypertarget{performance}{%
\section{Performance}\label{performance}}

\begin{itemize}
\tightlist
\item
  A major performance problem in our code is that small learning rates
  for the beta-booster cause processing time of the code to increase
  sharply. This remains to be solved.
\end{itemize}

@@

\hypertarget{further}{%
\section{Further}\label{further}}

\begin{itemize}
\item
  Allow user-input of the LocationScaleRegressionBoost-model in form of
  a dataframe, for example by allowing to pass an optional
  ``data''-argument as known from the lm-call. This may reduce proneness
  to input-errors.
\item
  Visualize the results of the boosted estimates
\item
  Document the extension of the R6-Class properly and inherit
  documentation of the LocationScaleRegression-Class.
\end{itemize}

\hypertarget{application}{%
\section{Application}\label{application}}

{[}Johannes: Munich Rent Data? - Identify Outliers{]}

\end{document}

### 2020-06-04 Reports/sample_report/1_REPORT.tex
write first report

\usepackage{graphicx,grffile}

@@

\title{1 Report ASP20 Boost}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{First Report}
\author{Johannes Strauß \and Levin Wiebelt \and Sebastian Aristizabal}
\date{28 Mai 2020}

@@

If you don't have Latex download latex for R with
\texttt{tinytex::install\_tinytex()}.

\hypertarget{write}{%
\section{Write}\label{write}}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# automatically create a bib database for R packages}
\NormalTok{knitr}\OperatorTok{::}\KeywordTok{write_bib}\NormalTok{(}\KeywordTok{c}\NormalTok{(}
  \KeywordTok{.packages}\NormalTok{(), }\StringTok{'R6'}\NormalTok{, }\StringTok{'knitr'}\NormalTok{, }\StringTok{'rmarkdown'}\NormalTok{, }\StringTok{'asp20model'}\NormalTok{, }\StringTok{'gamboostLSS'}\NormalTok{,}\StringTok{'mboost'}\NormalTok{, }\StringTok{'tidyverse'}  
\NormalTok{), }\StringTok{'packages.bib'}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{itemize}
\tightlist
\item
  The developement of this package takes place in the framework of the
  seminar ``Advanced Statisical Programming'' -\textgreater{} SS20
\item
  We expand the R6 class \emph{``LocationScaleRegression''} belonging to
  the \texttt{asp20model} package concieved specifficaly for this
  seminar (Riebl 2020)
\item
  Our aim is to develope an multi-faceted implementation of boosting for
  location and scale including component-wise boosting, use of
  cross-validation to determine the optimal stopping critera and useful
  options for visualization.
\end{itemize}

@@

JS: Performance Needs to be checked and maybe more outputs - messages?
Starting point of Beta and Gamma?

@@

@@

\begin{itemize}
\tightlist
\item
  boosting for multidimensional (\textgreater2) beta or gamma
\item
  component-wise boosting
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    allow for model creation via dataframe-input
  \item
    this may reduce errors due to unexpected inputs
  \end{itemize}
\item
  illustrate functions of package with real dataframe (= Vignette)
\item
  Visualization of results via connecting to plot team
\item
  functionality: predict
\item
  Include model diagnostics
\end{itemize}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-R-R6}{}%
Chang, Winston. 2019. \emph{R6: Encapsulated Classes with Reference
Semantics}. \url{https://CRAN.R-project.org/package=R6}.

@@

\leavevmode\hypertarget{ref-R-asp20model}{}%
Riebl, Hannes. 2020. \emph{Asp20model: An R6 Class for Location-Scale
Regression Models}.

>>>>

\usepackage{graphicx}

@@

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{1 Report ASP20 Boost}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{First Report}
\author{Johannes Strauß \and Levin Wiebelt \and Sebastian Aristizabal}
\date{03 June 2020}

@@

If you don't have Latex download latex for R with
\texttt{tinytex::install\_tinytex()}. To install pandoc on Mac, install
it via Terminal with \texttt{brew\ install\ pandoc-citeproc}. \#\# Write

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# automatically create a bib database for R packages}
\NormalTok{knitr}\OperatorTok{::}\KeywordTok{write\_bib}\NormalTok{(}\KeywordTok{c}\NormalTok{(}
  \KeywordTok{.packages}\NormalTok{(), }\StringTok{\textquotesingle{}R6\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}knitr\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}rmarkdown\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}asp20model\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}gamboostLSS\textquotesingle{}}\NormalTok{,}\StringTok{\textquotesingle{}mboost\textquotesingle{}}\NormalTok{, }\StringTok{\textquotesingle{}tidyverse\textquotesingle{}}  
\NormalTok{), }\StringTok{\textquotesingle{}packages.bib\textquotesingle{}}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{itemize}
\tightlist
\item
  The developement of this package takes place in the framework of the
  seminar ``Advanced Statisical Programming'' -\textgreater{} SS20
\item
  We extend the R6 class \emph{``LocationScaleRegression''} belonging to
  the \texttt{asp20model} package concieved specifficaly for this
  seminar (Riebl 2020)
\item
  Our aim is to develope an multi-faceted implementation of boosting for
  location and scale regression including component-wise boosting, use
  of cross-validation to determine the optimal stopping critera and
  useful options for visualization.
\end{itemize}

@@

Our package rely heavily on the asp20model package, any changes, or
errors in the code of the package might break the functionality.
Currently there is no validation of the given input, with respect to
heteroscedasticity.

The package also does not provide any auto optimization of the gradient
boost step-size, a wrongly picked step-size can lead to a deficient
performance, but an already implemented early stopping mechanism can
prevent an overfitting. An input check and cross-validation of the
step-size is on the to-do list. Also, the initial starting values for
Beta and Gamma can be further improved, see (Fahrmeir 2013, 219).
Further optimizations can be made to reduce the computation time, by
moving the calculation of the Projection Matrix for individual
covariates in the initialization section of the model. In regard to
User-friendliness of the package, we will try to implement further input
validation and try to write up with a comprehensive vignette containing
various examples in the next weeks.

@@

JS: During the learning process, we tried several methods to implement
the required functionality. At first, we started with a simple function
in a R-File. While looking at other online resources, we agreed on using
the asp20model packages as a requirement for our package and inheriting
the existing LocationScaleRegression class. Our
\texttt{LocationScaleRegressionBoost} R6 class implements a boosting
algorithm, two functions within the model for calculating the best
fitting covariate with respect to Gamma and Beta and has four new
variables which are set at the initialization of the model to reduce
computation time of the Projection Matrixes during the gradient boosting
(subject to change). The package provides the gradient boost function
for the \texttt{LocationScaleRegressionBoost} class, which uses the two
model functions to optimize stepwise the Beta and Gamma Vector.

@@

\begin{itemize}
\tightlist
\item
  boosting for multidimensional (\textgreater2) beta and gamma
\item
  component-wise boosting
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    allow for model creation via dataframe-input
  \item
    this may reduce errors due to unexpected inputs
  \end{itemize}
\item
  illustrate functions of package with real dataframe (= Vignette)
\item
  Visualization of results via connecting to plot team
  \texttt{asp20plot}
\item
  functionality: predict
\item
  Include model diagnostics
\end{itemize}

JS: check when run gradientboost, if model is
LocationScaleRegressionBoost class

@@

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-R-R6}{}%
Chang, Winston. 2019. \emph{R6: Encapsulated Classes with Reference
Semantics}. \url{https://CRAN.R-project.org/package=R6}.

@@

\leavevmode\hypertarget{ref-R-asp20model}{}%
Riebl, Hannes. 2020. \emph{Asp20model: An R6 Class for Location-Scale
Regression Models}.
\end{cslreferences}
